An integrated open population distance sampling approach for modelling age-structured populations

### 2024-01-05 Manuscript_Nilsen_Nater.qmd
Edits to introduction and methods

@@

Until recently, joint estimation of population dynamics and demography has relied mostly on data from marked individuals through application of open-population capture-mark-recapture models [@schaub2021]. While such methods can provide valuable information for both ecological research and management, collecting the necessary data is typically costly and logistically challenging to implement over large areas. Monitoring focusing on abundance trends over larger areas, on the other hand, are typically based on data from unmarked animals (ref). A common approach for such surveys is to structure data collection around the distance sampling (DS) method. DS has been used for estimating animal abundance in a wide range of contexts and for a variety of taxa [@buckland2015]. One reason for the method's popularity is that it does not require repeated visits to the same sites for estimating detection probability. This makes DS particularly useful for implementation in participatory monitoring programs, allowing stakeholders to take part in the data collection process.

The most common implementations of DS models have long used close-population formulations, and thus do not allow formulating a process model that projects abundance across years based on estimates of population growth rate ($\lambda$) or underlying demographic rates @buckland2015. In recent years, DS approaches have been extended in many ways, including applications that estimate changes in abundance over time in open populations via a hidden state model representing population dynamics [@moore2011; @sollmann2015]. This has greatly extended the potential of DS approaches for studying ecological dynamics across time and space. However, while these latter frameworks may allow to accurately quantify population changes, they typically provide little information on the drivers of these changes, i.e. the underlying vital rates. In fact, if the data does not contain information about the age (and/or sex-) structure of the surveyed population, there is no straightforward way to estimate demographic rates from such data. On the contrary, if age (and sex) of detected individuals can be determined, this information can be used to provide information on recruitment rates and survival probabilities. @Nilsen2018, for example, used a model based on harvest statistics and observations of population age structure to estimate population abundance and demographic rates without the need for any additional data from marked individuals.

@@

@@

Vital rates (survival probabilities $S$, recruitment rates $R$), detection parameters (half-normal detection parameters $\sigma$), and initial population densities can all be modelled as time-dependent in our framework. For both the tests with simulated data and the case study described below, we implemented log-normally distributed random year effects on all parameters except survival, which was set to be constant. In the case study, we additionally included an effect of rodent occupancy (see details below) on log recruitment rates, resulting in the following model:

@@

We assessed the model's overall performance and ability to estimate abundance, demographic rates, and detection parameters without bias by testing it on simulated data. We generated a total of 10 simulated datasets in five steps. First, we simulated 15-year time-series of survival and recruitment rates from biologically plausible values for averages of and -- in the case of recruitment -- among-year variation in demographic rates (survival was held constant across years). Second, we used the yearly demographic rate and realistic initial population densities to simulate stochastic population dynamics in 50 distinct sites. Third, we simulated the grouping of individuals in each site by first determining the expected number of groups in a site (based on the average group size of 5.6 from our ptarmigan data) and then distributing individuals among groups via multinomial trials. Fourth, we assigned a distance from transect line to each group and simulated the line transect survey in all 50 sites across 15 years. Finally, we simulated 5-year time-series of radio-telemetry data (= survival from one year to the next) for a subset of individuals (30 on average) using the simulated survival probabilities for each relevant year. We then fit the IDSM to each of the 10 simulated datasets three times, using distinct seeds for both simulating initial values and initiating and running the MCMC. Model implementation for simulated data tests was largely identical to that for real data and is described in detail below.

@@

For the model fits to simulated and real data we ran 3 and 4 MCMC chains with NIMBLE's standard samples for 500k and 100k iterations, respectively. 300k and 40k and thereof were discarded as burn-in prior to thinning with factors 5 and 20 , leaving us with 40k and 3k posterior samples per chain (total of 120k and 12k samples per run), repectively.

>>>>

<!-- Just a note: we seems to also still be missing quite a few references in the introduction -->

@@

Until recently, joint estimation of population dynamics and demography has relied mostly on data from marked individuals and associated open-population capture-mark-recapture models [@schaub2021]. While such methods can provide valuable information for both ecological research and management, collecting the necessary data is typically costly and logistically challenging to implement over large areas. Monitoring programmes focusing on abundance trends over larger areas, on the other hand, are typically based on data from unmarked animals (ref). A common approach for such surveys is to structure data collection around the distance sampling (DS) method. DS has been used for estimating animal abundance in a wide range of contexts and for a variety of taxa [@buckland2015]. One reason for the method's popularity is that it does not require repeated visits to the same sites for estimating detection probability. This makes DS particularly useful for implementation in participatory monitoring programs, allowing stakeholders to take part in the data collection process.

The most common implementations of DS models have long used closed-population formulations, and thus do not allow formulating a process model that projects abundance across years based on estimates of population growth rate ($\lambda$) or underlying demographic rates @buckland2015. In recent years, DS approaches have been extended in many ways, including applications that estimate changes in abundance over time in open populations via a hidden state model representing population dynamics [@moore2011; @sollmann2015]. This has greatly extended the potential of DS approaches for studying ecological dynamics across time and space. However, while these latter frameworks may allow to accurately quantify population changes, they typically provide little information on the drivers of these changes, i.e. the underlying vital rates. In fact, if the data does not contain information about the age (and/or sex-) structure of the surveyed population, there is no straightforward way to estimate demographic rates from such data. On the contrary, if age (and sex) of detected individuals can be determined, this information can be used to provide information on recruitment rates and survival probabilities. @Nilsen2018, for example, used a model based on harvest statistics and observations of population age structure to estimate population abundance and demographic rates without the need for any additional data from marked individuals.

@@

<!-- Note: missing ref below. Maybe good to "search" the whole text for "ref" to make sure we have replaced the placeholder with appropriate references everywhere -->

@@

Vital rates (survival probabilities $S$, recruitment rates $R$) and detection parameters (half-normal detection parameters $\sigma$) can all be modelled as time-dependent in our framework. For both the tests with simulated data and the case study described below, we implemented log-normally distributed random year effects on all parameters except survival, which was set to be constant. In the case study, we additionally included an effect of rodent occupancy (see details below) on log recruitment rates, resulting in the following model:

@@

We assessed the model's overall performance and ability to estimate abundance, demographic rates, and detection parameters without bias by testing it on simulated data. We generated a total of 10 simulated datasets in five steps. First, we simulated 15-year time-series of survival and recruitment rates from biologically plausible values for averages and -- in the case of recruitment -- among-year variation in demographic rates (survival was held constant across years). Second, we used the yearly demographic rate and realistic initial population densities to simulate stochastic population dynamics in 50 distinct sites. Third, we simulated the grouping of individuals in each site by first determining the expected number of groups in a site (based on the average group size of 5.6 from our ptarmigan data) and then distributing individuals among groups via multinomial trials. Fourth, we assigned a distance from transect line to each group and simulated the line transect survey in all 50 sites across 15 years. Finally, we simulated 5-year time-series of radio-telemetry data (= survival from one year to the next) for a subset of individuals (30 per year on average) using the simulated survival probabilities for each relevant year. We then fit the IDSM to each of the 10 simulated datasets three times, using distinct seeds for both simulating initial values and initiating and running the MCMC. Model implementation for simulated data tests was largely identical to that for real data and is described in detail below.

@@

For the model fits to simulated and real data we ran 3 and 4 MCMC chains with NIMBLE's standard samples for 500k and 100k iterations, respectively. 300k and 40k and thereof were discarded as burn-in prior to thinning with factors 5 and 20 , leaving us with 40k and 3k posterior samples per chain (total of 120k and 12k samples per run), respectively.
